---
title: "README"
author: "Ryan^3 + John"
date: "2024-03-29"
output: html_document
---

Group Name: 
**Ryan^3 + John**

Group Members:

- **Ryan Ledbetter**
- **Ryan Riebesehl**
- **Ryan Rockey**
- **John Nesnidal**


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## English Premier League Statistics

Our data encompasses the last 22 years of the English Premier League, the largest soccer league in the world, by revenue. This includes statistics about every game, including goals scored, shots taken, the teams involved, even betting odds for many of the games.

```{r}
prem <- read.csv("Datasets/final_dataset.csv")
head(prem, 3)
```

### Questions to be answered:

#### Trends over time:
  - How has scoring changed over the years?
  - How does scoring change across a full season?

#### Home vs. Away performance:
  - How do teams tend to perform in home games vs. away games?
  - Is there a measurable advantage to being at home?

#### Goal scoring analysis:
  - Which teams have scored the most/fewest goals?
  - Does having a greater % of shots on target correlate to winning more?
  
#### Betting trends:
  - How well do the betting odds predict the outcome of a game?
  
#### Predictive analysis:
  - Can we use the previous data to predict the outcome of the rest of the current premier league season?




