---
title: "README"
author: "Ryan^3 + John"
date: "2024-03-29"
output: html_document
---

Group Name: **Ryan\^3 + John**

Group Members:

-   **Ryan Ledbetter**
-   **Ryan Riebesehl**
-   **Ryan Rockey**
-   **John Nesnidal**

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(dplyr)
library(tidyr)
```

## English Premier League Statistics

Our data encompasses the last 22 years of the English Premier League, the largest soccer league in the world, by revenue. This includes statistics about every game, including goals scored, shots taken, the teams involved, even betting odds for many of the games.

```{r}
##prem <- read.csv(file.choose(), header = T)
prem <- read.csv("Datasets/final_dataset.csv")

head(prem, 3)
```

```{r}
##Missing Values:
missing_values <- colSums(is.na(prem))
print(missing_values)
```

```{r}
##Date-time Format:
prem$Date <- as.Date(prem$Date, format = "%d/%m/%y")

##Unique Values:
unique_teams <- unique(c(prem$HomeTeam, prem$AwayTeam))
print(unique_teams)

## Possible Unnecessay Columns:
columns_to_remove <- c("Div", "Time", "Referee") 
prem_clean <- prem[, !(names(prem) %in% columns_to_remove)]
```

## **Data Explanation:**

Our dataset consisted of many data points but these were not found across all of the csv files provided. The columns that were provided include: Our dataset has many fields, those include:

Div: (String) Representss the League Division\
Date: (Date) Match Date (dd/mm/yy)\
Time: (Time) Time of match (Military time)\
HomeTeam: (String) Home Team\
AwayTeam: (String) Away Team\
FTHG: (Int) Full Time Home Team Goals FTAG: (Int) Full Time Away Team Goals\
FTR: (H=Home Win, D=Draw, A=Away Win) Full Time Result\
HTHG: (Int) Half Time Home Team Goals\
HTAG: (Int) Half Time Away Team Goals\
HTR: (H=Home Win, D=Draw, A=Away Win) Half Time Result

and Match Statistics (where available)\
Attendance: (Int) Crowd Attendance\
Referee: (String) Match Referee\
HS: (Int) Home Team Shots\
AS: (Int) Away Team Shots\
HST: (Int) Home Team Shots on Target\
AST: (Int)Away Team Shots on Target\
HHW: (Int) Home Team Hit Woodwork\
AHW: (Int) Away Team Hit Woodwork\
HC: (Int) Home Team Corners\
AC: (Int) Away Team Corners\
HF: (Int) Home Team Fouls Committed\
AF: (Int) Away Team Fouls Committed\
HO: (Int) Home Team Offsides\
AO: (Int) Away Team Offsides\
HY: (Int) Home Team Yellow Cards\
AY: (Int) Away Team Yellow Cards\
HR: (Int) Home Team Red Cards\
AR: (Int) Away Team Red Cards HBP: (Int) Home Team Bookings Points (10 : yellow, 25 : red)\
ABP: (Int) Away Team Bookings Points (10 : yellow, 25 : red)

Additionally there were more betting odds representing multiple forms of betting such as

<details>

<summary>Additional Columns</summary>

```         
  Key to 1X2 (match) betting odds data:  
    B365H:   
    (Float) Bet365 home win odds  
    B365D:   
    (Float) Bet365 draw odds  
    B365A:   
    (Float) Bet365 away win odds  
    BSH:   
    (Float) Blue Square home win odds  
    BSD:   
    (Float) Blue Square draw odds  
    BSA:   
    (Float) Blue Square away win odds  
    BWH:   
    (Float) Bet&Win home win odds  
    BWD:   
    (Float) Bet&Win draw odds  
    BWA:   
    (Float) Bet&Win away win odds  
    GBH:   
    (Float) Gamebookers home win odds  
    GBD:   
    (Float) Gamebookers draw odds  
    GBA:   
    (Float) Gamebookers away win odds  
    IWH:   
    (Float) Interwetten home win odds  
    IWD:   
    (Float) Interwetten draw odds  
    IWA:   
    (Float) Interwetten away win odds  
    LBH:   
    (Float) Ladbrokes home win odds  
    LBD:   
    (Float) Ladbrokes draw odds  
    LBA:   
    (Float) Ladbrokes away win odds  
    PSH:   
    (Float) Pinnacle Sports home win odds  
    PSD:   
    (Float) Pinnacle Sports draw odds  
    PSA:   
    (Float) Pinnacle Sports away win odds  
    SOH:   
    (Float) Sporting Odds home win odds  
    SOD:   
    (Float) Sporting Odds draw odds  
    SOA:   
    (Float) Sporting Odds away win odds  
    SBH:   
    (Float) Sportingbet home win odds  
    SBD:   
    (Float) Sportingbet draw odds  
    SBA:   
    (Float) Sportingbet away win odds  
    SJH:   
    (Float) Stan James home win odds  
    SJD:   
    (Float) Stan James draw odds  
    SJA:   
    (Float) Stan James away win odds  
    SYH:   
    (Float) Stanleybet home win odds  
    SYD:   
    (Float) Stanleybet draw odds  
    SYA:   
    (Float) Stanleybet away win odds  
    VCH:   
    (Float) VC Bet home win odds  
    VCD:   
    (Float) VC Bet draw odds  
    VCA:   
    (Float) VC Bet away win odds  
    WHH:   
    (Float) William Hill home win odds  
    WHD:   
    (Float) William Hill draw odds  
    WHA:   
    (Float) William Hill away win odds  
      
    Bb1X2:    
    (Float) Number of BetBrain bookmakers used to calculate match odds averages and maximums  
    BbMxH:    
    (Float) Betbrain maximum home win odds  
    BbAvH:    
    (Float) Betbrain average home win odds  
    BbMxD:    
    (Float) Betbrain maximum draw odds  
    BbAvD:    
    (Float) Betbrain average draw win odds  
    BbMxA:    
    (Float) Betbrain maximum away win odds  
    BbAvA:    
    (Float) Betbrain average away win odds  
      
      
      
    Key to total goals betting odds:  
      
    BbOU:    
    (Float) Number of BetBrain bookmakers used to calculate over/under 2.5 goals (total goals) averages and maximums    
    BbMx>2.5:    
    (Float) Betbrain maximum over 2.5 goals  
    BbAv>2.5:    
    (Float) Betbrain average over 2.5 goals  
    BbMx<2.5:    
    (Float) Betbrain maximum under 2.5 goals  
    BbAv<2.5:    
    (Float) Betbrain average under 2.5 goals  
      
    GB>2.5:    
    (Float) Gamebookers over 2.5 goals  
    GB<2.5:    
    (Float) Gamebookers under 2.5 goals  
    B365>2.5:    
    (Float) Bet365 over 2.5 goals  
    B365<2.5:    
    (Float) Bet365 under 2.5 goals  
      
      
    Key to Asian handicap betting odds:  
      
    BbAH:    
    (Float) Number of BetBrain bookmakers used to Asian handicap averages and maximums  
    BbAHh:    
    (Float) Betbrain size of handicap (home team)  
    BbMxAHH:    
    (Float) Betbrain maximum Asian handicap home team odds  
    BbAvAHH:    
    (Float) Betbrain average Asian handicap home team odds  
    BbMxAHA:    
    (Float) Betbrain maximum Asian handicap away team odds  
    BbAvAHA:    
    (Float) Betbrain average Asian handicap away team odds  
      
    GBAHH:    
    (Float) Gamebookers Asian handicap home team odds  
    GBAHA:    
    (Float) Gamebookers Asian handicap away team odds  
    GBAH:    
    (Float) Gamebookers size of handicap (home team)  
    LBAHH:    
    (Float) Ladbrokes Asian handicap home team odds  
    LBAHA:    
    (Float) Ladbrokes Asian handicap away team odds  
    LBAH:    
    (Float) Ladbrokes size of handicap (home team)  
    B365AHH:    
    (Float) Bet365 Asian handicap home team odds  
    B365AHA:    
    (Float) Bet365 Asian handicap away team odds  
    B365AH:    
    (Float) Bet365 size of handicap (home team)  
```

</details>

## **Marginal Summaries:**

```{r}
##Summary Statistics:
summary(prem_clean)
```

```{r}
##Table of full time results:
outcome_freq <- table(prem_clean$FTR)
print(outcome_freq)
```

### Questions to be answered:

#### General Trends over time: Ryan Riebesehl

-   How has scoring changed over the years?
-   How does scoring change across a full season?
-   Is there a significant home advantage in the English Premier League over the past 22 years?

#### Home vs. Away performance: Ryan L

-   How do teams tend to perform in home games vs. away games?
-   Is there a measurable advantage to being at home?
-   Is there a disadvantage to playing as the away team?

#### Goal scoring analysis: Ryan Rockey

-   Which teams have scored the most/fewest goals?
-   Does having a greater % of shots on target correlate to winning more?
-   Is there a correlation between the number of shots taken by a team and the number of goals scored?

------------------------------------------------------------------------

#### [General Trends over time:]{.underline}

(1). How has Scoring Changed Over Time?

```{r}
library(dplyr)  
library(ggplot2)
library(tidyverse)

#Add column for season (year):
prem_clean <- prem_clean %>%
  mutate(Season = format(Date, "%Y"))

#Average goals per game for each season:
avg_goals_per_season <- prem_clean %>%
  group_by(Season) %>%
  summarize(
    avg_goals = mean(FTHG + FTAG)  #(home goals + away goals)
  )

# Line Graph of the average goals per season:
ggplot(avg_goals_per_season, aes(x = Season, y = avg_goals, group = 1)) +
  geom_line() +
  labs(title = "Average Goals Per Game Over the Years",
       x = "Season (Year)",
       y = "Average Goals Per Game") +
  theme_minimal()

```

\*Based on our line graph, the average goals per game were lower near the older years (2000-2009). While the average goals per game were higher in the more recent years (2010-2020). One reason why the average goals are increasing over the years could be due to multiple thrashings occurring between some teams and the rest of the league. Another reason could be that Premier League fans like a high goals-per-game ratio. Or from a strategic standpoint, there is a lack of defensive organisation.

\*[Resource]{.underline}:<https://theathletic.com/5292045/2024/02/24/why-so-many-goals-premier-league/>

\-

(2). How does scoring change across a full season?

```{r}
#Column for matchday:
prem_clean <- prem_clean %>%
  mutate(Matchday = as.integer(format(Date, "%j")) %/% 7 + 1)

#Average goals per game for each matchday:
avg_goals_per_matchday <- prem_clean %>%
  group_by(Matchday) %>%
  summarize(
    avg_goals = mean(FTHG + FTAG)  #(home goals + away goals)
  )

#Line graph of average goals per matchday:
ggplot(avg_goals_per_matchday, aes(x = Matchday, y = avg_goals, group = 1)) +
  geom_line() +
  labs(title = "Average Goals Per Game Across the Season",
       x = "Matchday",
       y = "Average Goals Per Game") +
  theme_minimal()

```

\*Based on our line graph, the average goals as the season progresses is the lowest at the start of the season an at the end of the season. Where the average goals is the highest at the mid-point of the season. This could be due to Premier League teams having the intensity of needing a reputation to intimidate other player in light of the championship playoffs at the end of the season.

\*[Resource]{.underline}:<https://theanalyst.com/na/2023/12/goals-galore-how-the-premier-league-is-producing-more-goals-than-ever-before/>

\-

(3). Is there a significant home advantage in the English Premier League over the past 22 years?

```{r}
#Precentages of home wins and away wins:
home_advantage <- prem_clean %>%
  group_by(Season) %>%
  summarize(
    home_win_percentage = mean(FTR == "H") * 100,
    away_win_percentage = mean(FTR == "NH") * 100,
  )
print(home_advantage)

#Chi-Square Test of Independe (Are these variables ralted?):
contingency_table <- table(prem_clean$FTR)
chi_square_test <- chisq.test(contingency_table)

print(chi_square_test)

#T-test compaing average goals scored by home and away teams (Is there a difference in the means between the variables?):
t_test <- t.test(prem_clean$FTHG, prem_clean$FTAG)

print(t_test)

#Line Graph Comparing home and away precentages over the years:
ggplot(home_advantage, aes(x = Season)) +
  geom_line(aes(y = home_win_percentage, color = "Home Wins", group = 1)) +
  geom_line(aes(y = away_win_percentage, color = "Away Wins", group = 2)) +
  labs(title = "Home/Away Advantage in the EPL Over the Past 22 Years",
       x = "Season (Year)",
       y = "Percentage") +
  scale_color_manual(values = c("Home Wins" = "blue", "Away Wins" = "red")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

\*Based on our line graph and percentages calculated, the win percentages tended to be higher for teams when they were away versus being on their home field. Additionally, according to our ran Chi-squared test and t-test, there is very strong evidence that the home and away variables are related to each other and their is a significant difference between them. While this quite a surprising result, some factors that could have contributed to this result could be due to a new environment (field) for the away players, the absence of fans during bad weather, or even that many people are watching the broadcasts of the games with their technological devices.

\*[Resource]{.underline}: <https://www.skysports.com/football/news/11661/12079338/premier-league-goals-and-away-wins-at-record-highs-has-home-advantage-been-wiped-out>

------------------------------------------------------------------------

#### [Home vs. Away performance]{.underline}:

```{r}
# List of CSV files
file_list <- c("Datasets/2000-01.csv",
"Datasets/2001-02.csv",
"Datasets/2002-03.csv",
"Datasets/2003-04.csv",
"Datasets/2004-05.csv",
"Datasets/2005-06.csv",
"Datasets/2006-07.csv",
"Datasets/2007-08.csv",
"Datasets/2008-09.csv",
"Datasets/2009-10.csv",
"Datasets/2010-11.csv",
"Datasets/2011-12.csv",
"Datasets/2012-13.csv",
"Datasets/2013-14.csv",
"Datasets/2014-15.csv",
"Datasets/2015-16.csv",
"Datasets/2016-17.csv",
"Datasets/2017-18.csv",
"Datasets/2018-19.csv",
"Datasets/2019-20.csv",
"Datasets/2020-2021.csv",
"Datasets/2021-2022.csv")

# Read CSV files into a list of data frames
data_frames <- lapply(file_list, read.csv)

# Identify common columns
common_columns <- Reduce(intersect, lapply(data_frames, colnames))

# Subset data frames to keep only common columns
data_frames_subset <- lapply(data_frames, function(df) df[, common_columns])

# Merge data frames by rows
merged_data <- do.call(rbind, data_frames_subset)

# Write merged data frame to CSV
write.csv(merged_data, "merged_file.csv", row.names = FALSE)
```

(1) How do teams tend to perform in home games vs. away games?

```{r}
avg_home_final_score <- mean(merged_data$FTHG)
avg_away_final_score <- mean(merged_data$FTAG)
avg_home_shots <- mean(merged_data$HST)
avg_away_shots <- mean(merged_data$AST)
avg_home_corners <- mean(merged_data$HC)
avg_away_corners <- mean(merged_data$AC)

final_score_percent = avg_home_final_score/avg_away_final_score
shots_percent = avg_home_shots/avg_away_shots
corners_percent = avg_home_corners/avg_away_corners

cat("The average final score per game by the home team is: ", avg_home_final_score, "\n") 
cat("The average fianl score per game by the away team is: ", avg_away_final_score, "\n")
cat("The average shots on goal by the home team is: ", avg_home_shots, "\n") 
cat("The average shots on goal by the away team is: ", avg_away_shots, "\n")
cat("The average amount of corner kicks by the home team is: ", avg_home_corners, "\n") 
cat("The average amount of corner kicks by the away team is: ", avg_away_corners, "\n\n")

cat("The home team scores ", final_score_percent, " times more points.\n")
cat("The home team shoots on goal ", shots_percent, " times more.\n")
cat("The home team gets ", corners_percent, " times more corner kicks.\n")
```

(2) Is there a measurable advantage to being at home?

```{r}
home_wins <- table(merged_data$FTR == "H")
home_wins <- as.data.frame(home_wins)[["Freq"]][2]

away_wins <- table(merged_data$FTR == "A")
away_wins <- as.data.frame(away_wins)[["Freq"]][2]

draws <- table(merged_data$FTR == "D")
draws <- as.data.frame(draws)[["Freq"]][2]

FTR_counts = table(merged_data$FTR)
percentages <- round(prop.table(FTR_counts) * 100, 1)

pie_chart <- ggplot(data = NULL, aes(x = "", y = FTR_counts, fill = names(FTR_counts))) +
  geom_bar(stat = "identity") +
  coord_polar("y", start=0) + 
  labs(title = "Distributions of Final Results for EPL", fill = "Result of Game") +
  theme_minimal() +
  theme(legend.position = "right", axis.text = element_blank(),
        axis.ticks = element_blank(),
        panel.grid  = element_blank(),
        axis.title = element_blank()) +
   scale_fill_manual(values = c("H" = "forestgreen", "A" = "maroon", "D" = "navy"),
                    labels = c(paste("Home Win", " (", percentages["H"], "%)"),
                               paste("Away Win", " (", percentages["A"], "%)"),
                               paste("Draw", " (", percentages["D"], "%)")))

# Display the pie chart
print(pie_chart)

cat("In our data set, of the 8020 matches,", home_wins, "were won by the home team, while ", away_wins, "matches were won by the away team. The amount of matches that ended in a draw was", draws, ".\n")
```

By looking at the data from our past two questions, it does appear that there is an advantage to being the home team. By playing at your own field, you seem to take more shots on goal, get more opportunities for corner kicks, and overall win more matches. This could all be due to a high amount of experience and practice on that field as well as having a crowd that is likely biased towards the home team. Competeing is a mental activity as well as physical so having a crowd there to keep you going could make up a part of the difference. We have seen that there is an advantage to being the team who gets to play at their own field, but could there also be a disadvantage to being the team that is away? We wanted to look at how likely you are to get called for a foul.

(3) Are teams more likely to get carded when away?

```{r}
avg_home_yellow <- mean(merged_data$HY)
avg_away_yellow <- mean(merged_data$AY)
avg_home_red <- mean(merged_data$HR)
avg_away_red <- mean(merged_data$AR)

avg_home_cards = avg_home_yellow + avg_home_red
avg_away_cards = avg_away_yellow + avg_away_red
card_percent = avg_away_cards/avg_home_cards

cat("The average amounts of cards per game recevied by the home team is: ", avg_home_cards, "\n") 
cat("The average amounts of cards per game recevied by the away team is: ", avg_away_cards, "\n")
cat("The away team is ", card_percent, " times more likely to receive a card.")

data <- data.frame(
  variable = c("Avg. Home Yellow Cards", "Avg. Away Yellow Cards", "Avg. Home Red Cards", "Avg. Away Red Cards"),
  average = c(avg_home_yellow, avg_away_yellow, avg_home_red, avg_away_red) 
)

ggplot(data, aes(x = variable, y = average, fill = variable)) +
  geom_bar(stat = "identity") +
  labs(title = "Average Count of Cards Received by Home and Away Teams",
       x = "Cards Received by Home and Away",
       y = "Average Cards Received per Game") +
  scale_fill_manual(values = c("Avg. Home Yellow Cards" = "yellow", 
                                "Avg. Away Yellow Cards" = "yellow", 
                                "Avg. Home Red Cards" = "red", 
                                "Avg. Away Red Cards" = "red")) +
  theme_dark() +  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

Like all sports, in soccer there are penalties. If a player commits a foul, they will receive a yellow or red card. This is from the discretion of the referee at the game. This brought up the idea that playing an away game may bias how often teams get carded. From our data we found that by being the away team, you are approximately 1.25 times more likely to receive a card. This could be due to bias but could also be explained for reasons such as a team playing more aggressively when they are away or teams not wanting to be fouled out at a home game.

\#### [Goal Scoring Analysis]{.underline}:

-   Which teams have scored the most/fewest goals?

```{r}
# Get unique teams from both HomeTeam and AwayTeam columns
unique_teams <- unique(c(merged_data$HomeTeam, merged_data$AwayTeam))

# Initialize empty vectors to store the total home and away goals
home_goals <- numeric(length(unique_teams))
away_goals <- numeric(length(unique_teams))

# Loop through each unique team
for (i in 1:length(unique_teams)) {
  team <- unique_teams[i]
  home_goals[i] <- sum(merged_data$FTHG[merged_data$HomeTeam == team])
  away_goals[i] <- sum(merged_data$FTHG[merged_data$AwayTeam == team])
}

# Create dataframe with team names and total goals
goals_scored_by_team <- data.frame(
  team = unique_teams,
  home_goals = home_goals,
  away_goals = away_goals
)

# Create total_goals column
goals_scored_by_team$total_goals = goals_scored_by_team$home_goals + goals_scored_by_team$away_goals

# Create most goals dataframe
most_goals_scored_by_team <- goals_scored_by_team %>%
  arrange(desc(total_goals)) %>%
  slice_head(n = 10)

# Create fewest goals dataframe
fewest_goals_scored_by_team <- goals_scored_by_team %>%
  arrange(total_goals) %>%
  slice_head(n = 10)

# Create plot of the most goals and fewest goals
plot1 <- ggplot(most_goals_scored_by_team, aes(x = reorder(team, -total_goals), y = total_goals)) + 
  geom_bar(stat = "identity") + 
  labs(x = "Team", y = "Total Goals") + 
  theme_minimal(base_size=9)

plot2 <- ggplot(fewest_goals_scored_by_team, aes(x = reorder(team, -total_goals), y = total_goals)) + 
  geom_bar(stat = "identity") + 
  labs(x = "Team", y = "Total Goals",) + 
  theme_minimal(base_size=9)

# Get the maximum value from both datasets
max_value <- max(max(most_goals_scored_by_team$total_goals), max(fewest_goals_scored_by_team$total_goals))

# Set the same y-axis limits for both plots
plot1 <- plot1 + ylim(0, max_value)
plot2 <- plot2 + ylim(0, max_value)

plot1
plot2
```

-   Does having a greater % of shots on target correlate to winning more?

```{r}
# Calculate on-target percentage for home and away teams
merged_data2 <- merged_data %>%
  mutate(HomeOnTargetPercent = (HST / HS) * 100,
         AwayOnTargetPercent = (AST / AS) * 100,
         HomeResult = ifelse(FTR == "H", "Win", ifelse(FTR == "A", "Loss", "Draw")),
         AwayResult = ifelse(FTR == "A", "Win", ifelse(FTR == "H", "Loss", "Draw")))

# Combine on-target percentage data for both home and away teams
on_target_data <- rbind(
  data.frame(OnTargetPercent = merged_data2$HomeOnTargetPercent, Result = merged_data2$HomeResult),
  data.frame(OnTargetPercent = merged_data2$AwayOnTargetPercent, Result = merged_data2$AwayResult)
)

# Plot histogram with combined data
ggplot(on_target_data, aes(x = OnTargetPercent, fill = Result)) +
  geom_histogram(bins = 10, binwidth = 10, position = "stack", alpha = 0.6) +
  scale_fill_manual(values = c("yellow", "red", "green")) +  # Specify colors for wins and losses
  labs(x = "On Target Percentage", y = "Frequency", title = "Histogram of On Target Percentage") + xlim(0,100) +
  theme_minimal()
```

-   Is there a correlation between the number of shots taken by a team and the number of goals scored?

```{r}
# Plot shots taken
ggplot(merged_data2, aes(x = HST, y = FTHG, fill = FTR)) +
  geom_point(shape = 21, size = 3, alpha = 0.7) +
  scale_fill_manual(values = c("red", "yellow", "green")) +
  labs(x = "Shots Taken", y = "Goals Scored", title = "Relationship between Shots Taken and Goals Scored for Home Team") +
  theme_minimal()

# Plot scatter plot with different colors for result
ggplot(merged_data2, aes(x = AST, y = FTAG, fill = FTR)) +
  geom_point(shape = 21, size = 3, alpha = 0.7) +
  scale_fill_manual(values = c("green", "yellow", "red")) +
  labs(x = "Shots Taken", y = "Goals Scored", title = "Relationship between Shots Taken and Goals Scored for Away Team") +
  theme_minimal()
```
